The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-08-13 script/en/chapters/primes.tex
fixed math mode in headings resulting in broken PDF bookmarks with \texorpdfstring; fixed float warnings by replacing placement [h] with [ht]; fixed warning about PDF version 1.5 by changing it to 1.4 in coverv1_*.pdf

   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
\begin{table}[h]
\begin{center}
\begin{tabular}{|cccc|}
\hline
	{\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Who} \\
\hline

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{MersenneNumbers02}{}
\subsection{Mersenne numbers $f(n) = 2^n - 1$  \quad for $ n $ prime}
    \index{Prime number!Mersenne} \index{Mersenne!prime number}
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems
    to deliver relatively large prime numbers but - as for $n=11$ 
    [$f(n)=2,047$] - it is repeatedly the case that the result even with
    prime exponents is \hyperlink{Mer-nums-not-always-prim}{not} prime.\\
    Today, all the Mersenne primes having less than around 4,000,000 digits
    are known (\hyperlink{M-39}{M-39}\index{Mersenne!prime number!M-39}):
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}
\end{itemize}      % be_2005_UPDATEN_if-new-mersenne-prime-appears

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection
   [Generalized Mersenne numbers $f(k,n) = k \cdot 2^n \pm 1$ / Proth numbers]
   {Generalized Mersenne numbers
   $f(k,n) = k \cdot 2^n \pm 1 $  for $ n $ prime and $ k $ small prime
    / Proth numbers\footnotemark}
    \footnotetext{%
       Their names come from the French farmer FranÁois Proth (1852-1879).
       More famous as the Proth primes is the related Sierpinski 
       problem\index{Number!Sierpinski}: Find all numbers k, so that
       $k*2^n+1$ is composite for all n > 0.
       See table \ref{L_n_Largest_Known-Primes}.
    }
   \label{generalized-mersenne-no1}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{generalizedMersennenumbers}{}
\subsection
   [Generalized Mersenne numbers $ f(b,n) = b^n \pm 1 $ / Cunningham project]
   {Generalized Mersenne numbers 
   $ f(b,n) = b^n \pm 1 $ / The Cunningham project}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{FermatNumbers02}{}
\subsection[Fermat numbers $f(n) = 2^{2^n} + 1$]
              {Fermat numbers\footnotemark~$f(n) = 2^{2^n} + 1$} 
    \footnotetext{%
       The Fermat prime numbers play a role in circle division.
       As proven by Gauss\index{Gauss, Carl Friedrich} a regular $p$-edge
       can only be constructed with the use of a pair of compasses and a 
       ruler, when $p$ is a Fermat prime number.
    }
    \label{L-FermatNumbers02}
    \index{Prime number!Fermat} \index{Fermat!prime number}       
    As mentioned \hyperlink{FermatNumbers01}{above} in 
    chapter~\ref{FermatNumbers01}, Fermat wrote to Mersenne regarding 
    his assumption, that all numbers of this type are primes.
    This assumption was disproved by Euler (1732). The prime $641$ divides $f(5)$\footnote{%
    Surprisingly this number can easily be found by using Fermat's theorem (see e.g. \cite[p. 176]{pr:Scheid1994})
    }. 
%    Surprisingly he would have been able obtain a positive result using
%    the negative prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
f(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
f(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
f(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
f(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
f(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = \mbox{65,537} &  \mapsto {\rm ~prime}  \\
f(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = \mbox{4,294,967,297} = 641 \cdot \mbox{6,700,417} &  \mapsto {\rm ~NOT~prime} ! \\
f(6) = 2^{2^6} + 1  = 2^{64} + 1 &  = \mbox{18,446,744,073,709,551,617} \\
                                 &  = \mbox{274,177} \cdot \mbox{67,280,421,310,721} & \mapsto {\rm ~NOT~prime} !\\
f(7) = 2^{2^7} + 1  = 2^{128} + 1 & = \mbox{(see page~\pageref{F7Morrison})}  
																 & \mapsto {\rm ~NOT~prime} !

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{generalizedFermatprimes}{}
\subsection[Generalized Fermat numbers $f(b,n) = b^{2^n} + 1$]
              {Generalized Fermat numbers\footnotemark~$f(b,n) = b^{2^n} + 1$}
    \footnotetext{%
      The base of this power is no longer restricted to 2.
      Even more generic would be:  $f(b,c,n) = b^{c^n} \pm 1$
    }
\label{generalized-fermat}
\index{Fermat!number!generalized} 
    Generalized Fermat numbers are more numerous than Mersenne numbers of a
    equal size and many of them are waiting to be discovered to fill the
    big gaps between the Mersenne primes already found or still undiscovered.
    Progress in number theory made it possible that numbers, where the
    representation is not limited to the base 2, can be tested at almost
    the same speed than a Mersenne number.

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection
    [Idea based on Euclid's proof $p_1 \cdot p_2 \cdots p_n +1$]
    {Idea based on Euclid's proof $p_1 \cdot p_2 \cdots p_n +1$}
This idea is based on \hyperlink{euclid}{Euclid's proof} that there are
infinite many prime numbers.
$$
\begin{array}{lll}
2{\cdot}3 +1 &      = 7 &          \mapsto {\rm ~prime} \\
2{\cdot}3{\cdot}5 +1 &      = 31    &      \mapsto {\rm ~prime} \\
2{\cdot}3{\cdot}5{\cdot}7 +1 &      = 211   &      \mapsto {\rm ~prime} \\
2{\cdot}3{\cdots}11 +1 &        = 2,311  &      \mapsto {\rm ~prime} \\
2\cdot3 \cdots 13 +1 &  = 59 \cdot 509 &    \mapsto {\rm ~NOT~prime} ! \\
2\cdot3 \cdots 17 +1 &  = 19 \cdot 97 \cdot 277 &   \mapsto {\rm ~NOT~prime} !
\\
\end{array}
$$

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection{As above but $-1$ except $+1$: $p_1 \cdot p_2 \cdots p_n -1$}
$$
\begin{array}{lll}
2\cdot 3 -1     &   = 5 &   \mapsto {\rm ~prime} \\
2\cdot 3 \cdot  5  -1   &   = 29 &  \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 7  -1   &   = 11 \cdot 19 & \mapsto {\rm ~NOT~prime} ! \\
2\cdot 3 \cdots 11 -1   &   = 2,309 &    \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 13 -1   &   = 30,029 &    \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 17 -1    &  = 61 \cdot 8,369 &   \mapsto {\rm ~NOT~prime!}
\end{array} 
$$

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection[Euclidean numbers $e_n = e_0 \cdot e_1 \cdots e_{n-1} + 1$]
              {Euclidean numbers $e_n = e_0 \cdot e_1 \cdots e_{n-1} + 1$
	       with $ n \geq 1 $ and $ e_0 := 1 $}
%	       with $n$ greater than or equal to $1$ and $e_0 := 1$}
    \index{Euclidean number}
    $e_{n-1}$ is not the $(n-1)$th prime number, but the number 
    previously found here.
    Unfortunately this formula is not open but recursive.
    The sequence starts with

@@

% --------------------------------------------------------------------------
% \pagebreak %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vskip +10 pt
\subsection{$f(n) = n^2 + n + 41$}
\label{L-Polynomfunktion01-41}
   This sequence starts off very {\em promisingly},  
   but is far from being a proof.

@@

% --------------------------------------------------------------------------
\pagebreak %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%   be_2005
% \vskip +10 pt
\subsection{$f(n) = n^2 - 79 \cdot n + 1,601$}
\label{L-Polynomfunktion02-1601}   \label{SrcPrimArith1}
   This function\footnote{%
   See \hyperlink{Primes_Appendix_A}{Appendix A of this chapter} for the source code
   to compute the table using Sage\index{Sage}.}  %Pari-GP.\index{Pari-GP}}
   delivers prime numbers for all values from $n=0$ to $n=79$.
Unfortunately $f(80) = 1,681 = 11 \cdot 151$ is not a prime number. To this
date, no function has been found that delivers more prime numbers in a row. On
the other hand, each prime occurs twice (first in the decreasing then in the
increasing sequence), which means that the algorithm delivers a total of 40
different prime values (these are the same ones as delivered by the function
in chapter~\ref{L-Polynomfunktion01-41})\footnote{%
   Another quadratic polynom, which delivers these primes, is: 
   $f(n) = n^2 - 9 \cdot n + 61$.\\
   Among the first 1000 sequence elements more than 50\% are prime
   (See \hyperlink{Primes_Appendix_A}{Appendix A of this chapter}).}.
$$
\begin{array}{|ll||ll|}
\hline
f(0)  = 1.601 & \mapsto {\rm ~prim}   &   f(26) =   223 & \mapsto {\rm ~prim} \\
f(1)  = 1.523 & \mapsto {\rm ~prim}   &   f(27) =   197 & \mapsto {\rm ~prim} \\
f(2)  = 1.447 & \mapsto {\rm ~prim}   &   f(28) =   173 & \mapsto {\rm ~prim} \\
f(3)  = 1.373 & \mapsto {\rm ~prim}   &   f(29) =   151 & \mapsto {\rm ~prim} \\
f(4)  = 1.301 & \mapsto {\rm ~prim}   &   f(30) =   131 & \mapsto {\rm ~prim} \\
f(5)  = 1.231 & \mapsto {\rm ~prim}   &   f(31) =   113 & \mapsto {\rm ~prim} \\
f(6)  = 1.163 & \mapsto {\rm ~prim}   &   f(32) =    97 & \mapsto {\rm ~prim} \\
f(7)  = 1.097 & \mapsto {\rm ~prim}   &   f(33) =    83 & \mapsto {\rm ~prim} \\
f(8)  = 1.033 & \mapsto {\rm ~prim}   &   f(34) =    71 & \mapsto {\rm ~prim} \\
f(9)  =   971 & \mapsto {\rm ~prim}   &   f(35) =    61 & \mapsto {\rm ~prim} \\
f(10) =   911 & \mapsto {\rm ~prim}   &   f(36) =    53 & \mapsto {\rm ~prim} \\
f(11) =   853 & \mapsto {\rm ~prim}   &   f(37) =    47 & \mapsto {\rm ~prim} \\
f(12) =   797 & \mapsto {\rm ~prim}   &   f(38) =    43 & \mapsto {\rm ~prim} \\
f(13) =   743 & \mapsto {\rm ~prim}   &   f(39) =    41 & \mapsto {\rm ~prim} \\
f(14) =   691 & \mapsto {\rm ~prim}   &   f(40) =    41 & \mapsto {\rm ~prim} \\              
f(15) =   641 & \mapsto {\rm ~prim}   &   f(41) =    43 & \mapsto {\rm ~prim} \\              
f(16) =   593 & \mapsto {\rm ~prim}   &   f(42) =    47 & \mapsto {\rm ~prim} \\              
f(17) =   547 & \mapsto {\rm ~prim}   &   f(43) =    53 & \mapsto {\rm ~prim} \\              
f(18) =   503 & \mapsto {\rm ~prim}   &   \cdots  &  \\                                       
f(19) =   461 & \mapsto {\rm ~prim}   &   f(77) = 1.447 & \mapsto {\rm ~prim} \\             
f(20) =   421 & \mapsto {\rm ~prim}   &   f(78) = 1.523 & \mapsto {\rm ~prim} \\             
f(21) =   383 & \mapsto {\rm ~prim}   &   f(79) = 1.601 & \mapsto {\rm ~prim} \\             
f(22) =   347 & \mapsto {\rm ~prim}   &   f(80) = 41 \cdot 41 & \mapsto {\rm ~NOT~prim!} \\
f(21) =   383 & \mapsto {\rm ~prim}   &   f(81) = 41 \cdot 43 & \mapsto {\rm ~NOT~prim!} \\ 
f(22) =   347 & \mapsto {\rm ~prim}   &   f(82) = 1.847 & \mapsto {\rm ~prim} \\             
f(23) =   313 & \mapsto {\rm ~prim}   &   f(83) = 1.933 & \mapsto {\rm ~prim} \\              
f(24) =   281 & \mapsto {\rm ~prim}   &   f(84) = 43 \cdot 47 &  \mapsto {\rm ~NOT~prim!} \\
f(25) =   251 & \mapsto {\rm ~prim}   &   & \\
\hline
\end{array} 
$$

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection[Polynomial functions 
    $f(x) = a_n x^n + a_{n-1}x^{n-1} + \cdots + a_1 x^1 + a_0$]
    {Polynomial functions\index{Polynomial} 
    $f(x) = a_n x^n + a_{n-1}x^{n-1} + \cdots + a_1 x^1 + a_0$ ~ 
    ($a_i$ in ${\mathbb Z}$, $n \geq 1$)}\index{Polynomial}

@@

     \begin{table}[h]
     %\begin{center}
     \hspace*{1cm}\begin{tabular}{|r|r|r|r|l|}
     \hline
     Elements & First element & Distance & When & Discovered by \\ \hline
     22 &  11,410,337,850,553 &   4,609,098,694,200 & 1993 & Paul A. Pritchard,\\
        &                     &                     &      & Andrew Moran,\\
        &                     &                     &      & Anthony Thyssen\\
        &                     &                     &      &\\

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|l|l||l|l||l|l|}\hline
\multicolumn{2}{|l||}{Ten-sized intervals} & \multicolumn{2}{l||}{Hundred-sized
intervals} & \multicolumn{2}{l|}{Thousand-sized intervals} \\ \hline
Interval  &     Number &    Interval  & Number &  Interval  & Number\\ \hline
\hline
1-10     &       4     &     1-100   &     25  &     1-1000     &    168 \\
11-20    &       4     &     101-200 &     21  &     1001-2000  &    135 \\
21-30    &       2     &     201-300 &     16  &     2001-3000  &    127  \\
31-40    &       2     &     301-400 &     16  &     3001-4000  &    120 \\
41-50    &       3     &     401-500 &     17  &     4001-5000  &    119 \\
51-60    &       2     &     501-600 &     14  &     5001-6000  &    114 \\
61-70    &       2     &     601-700 &     16  &     6001-7000  &    117 \\
71-80    &       3     &     701-800 &     14  &     7001-8000  &    107 \\
81-90    &       2     &     801-900 &     15  &     8001-9000  &    110 \\
91-100   &       1     &     901-1000 &     14 &      9001-10000 &    112 \\
\hline
\end{tabular}
\caption{How many primes exist within the first intervals of tens?}
\end{center}
\end{table}
\vskip +20 pt

\begin{table}[h]
\begin{center}
\begin{tabular}{|l|l|l|}\hline
Interval & Number & Average number per 1000 \\ \hline
1 - 10,000          &    1,229        &    122.900 \\
1 - 100,000         &    9,592        &    95.920 \\
1 - 1,000,000       &    78,498      &    78.498 \\
1 - 10,000,000       &   664,579     &    66.458 \\
1 - 100,000,000      &   5,761,455   &    57.615 \\
1 - 1,000,000,000    &   50,847,534  &    50.848 \\
1 - 10,000,000,000   &   455,052,512 &    45.505 \\ \hline
\end{tabular}
\caption{How many primes exist within the first intervals of dimensions?}
\end{center}
\end{table}
\vskip +6 pt

% --------------------------------------------------------------------------
\newpage
\hypertarget{ntePrimzahl}{}
\section{Appendix: Indexing prime numbers ($n$-th prime number)} 
\label{s:ntePrimzahl}
\vskip +10 pt

>>>>

   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
\begin{table}[ht]
\begin{center}
\begin{tabular}{|cccc|}
\hline
	{\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Who} \\
\hline

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{MersenneNumbers02}{}
\subsection{Mersenne numbers \texorpdfstring{$f(n) = 2^n - 1$  \quad for $ n $ prime}
	   			      {f(n) = 2\^{}n - 1 for n prime}}
    \index{Prime number!Mersenne} \index{Mersenne!prime number}
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems
    to deliver relatively large prime numbers but - as for $n=11$ 
    [$f(n)=2,047$] - it is repeatedly the case that the result even with
    prime exponents is \hyperlink{Mer-nums-not-always-prim}{not} prime.\\
    Today, all the Mersenne primes having less than around 4,000,000 digits
    are known (\hyperlink{M-39}{M-39}\index{Mersenne!prime number!M-39}):
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}
\end{itemize}      % be_2005_UPDATEN_if-new-mersenne-prime-appears

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection
   [Generalized Mersenne numbers \texorpdfstring{$f(k,n) = k \cdot 2^n \pm 1$}{f(k,n) = k .... 2\^{}n +- 1} / Proth numbers]
   {Generalized Mersenne numbers
   $f(k,n) = k \cdot 2^n \pm 1 $  for $ n $ prime and $ k $ small prime
    / Proth numbers\footnotemark}
    \footnotetext{%
       Their names come from the French farmer FranÁois Proth (1852-1879).
       More famous as the Proth primes is the related Sierpinski 
       problem\index{Number!Sierpinski}: Find all numbers k, so that
       $k*2^n+1$ is composite for all n > 0.
       See table \ref{L_n_Largest_Known-Primes}.
    }
   \label{generalized-mersenne-no1}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{generalizedMersennenumbers}{}
\subsection
   [Generalized Mersenne numbers \texorpdfstring{$ f(b,n) = b^n \pm 1 $}{f(b,n) = b\^{}n +- 1} / 
    Cunningham project]
   {Generalized Mersenne numbers 
   $ f(b,n) = b^n \pm 1 $ / The Cunningham project}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{FermatNumbers02}{}
\subsection[Fermat numbers \texorpdfstring{$f(n) = 2^{2^n} + 1$}{f(n) = 2\^{}2\^{}n + 1}]
              {Fermat numbers\footnotemark~$f(n) = 2^{2^n} + 1$} 
    \footnotetext{%
       The Fermat prime numbers play a role in circle division.
       As proven by Gauss\index{Gauss, Carl Friedrich} a regular $p$-edge
       can only be constructed with the use of a pair of compasses and a 
       ruler, when $p$ is a Fermat prime number.
    }
    \label{L-FermatNumbers02}
    \index{Prime number!Fermat} \index{Fermat!prime number}       
    As mentioned \hyperlink{FermatNumbers01}{above} in 
    chapter~\ref{FermatNumbers01}, Fermat wrote to Mersenne regarding 
    his assumption, that all numbers of this type are primes.
    This assumption was disproved by Euler (1732). The prime $641$ divides $f(5)$\footnote{%
    Surprisingly this number can easily be found by using Fermat's theorem (see e.g. \cite[p. 176]{pr:Scheid1994})
    }. 
%    Surprisingly he would have been able obtain a positive result using
%    the negative prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
f(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
f(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
f(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
f(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
f(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = \mbox{65,537} &  \mapsto {\rm ~prime}  \\
f(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = \mbox{4,294,967,297} = 641 \cdot \mbox{6,700,417} &  \mapsto {\rm ~NOT~prime} ! \\
f(6) = 2^{2^6} + 1  = 2^{64} + 1 &  = \mbox{18,446,744,073,709,551,617} \\
                                 &  = \mbox{274,177} \cdot \mbox{67,280,421,310,721} & \mapsto {\rm ~NOT~prime} !\\
f(7) = 2^{2^7} + 1  = 2^{128} + 1 & = \mbox{(see page~\pageref{F7Morrison})}  
																 & \mapsto {\rm ~NOT~prime} !

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{generalizedFermatprimes}{}
\subsection[Generalized Fermat numbers \texorpdfstring{$f(b,n) = b^{2^n} + 1$}{f(b,n) = b\^{}2\^{}n + 1}]
              {Generalized Fermat numbers\footnotemark~$f(b,n) = b^{2^n} + 1$}
    \footnotetext{%
      The base of this power is no longer restricted to 2.
      Even more generic would be:  $f(b,c,n) = b^{c^n} \pm 1$
    }
\label{generalized-fermat}
\index{Fermat!number!generalized} 
    Generalized Fermat numbers are more numerous than Mersenne numbers of a
    equal size and many of them are waiting to be discovered to fill the
    big gaps between the Mersenne primes already found or still undiscovered.
    Progress in number theory made it possible that numbers, where the
    representation is not limited to the base 2, can be tested at almost
    the same speed than a Mersenne number.

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection
    [Idea based on Euclid's proof \texorpdfstring{$p_1 \cdot p_2 \cdots p_n +1$}{p\_1 ... p\_2 ... p\_n +1}]
    {Idea based on Euclid's proof $p_1 \cdot p_2 \cdots p_n +1$}
This idea is based on \hyperlink{euclid}{Euclid's proof} that there are
infinite many prime numbers.
$$
\begin{array}{lll}
2{\cdot}3 +1 &      = 7 &          \mapsto {\rm ~prime} \\
2{\cdot}3{\cdot}5 +1 &      = 31    &      \mapsto {\rm ~prime} \\
2{\cdot}3{\cdot}5{\cdot}7 +1 &      = 211   &      \mapsto {\rm ~prime} \\
2{\cdot}3{\cdots}11 +1 &        = 2,311  &      \mapsto {\rm ~prime} \\
2\cdot3 \cdots 13 +1 &  = 59 \cdot 509 &    \mapsto {\rm ~NOT~prime} ! \\
2\cdot3 \cdots 17 +1 &  = 19 \cdot 97 \cdot 277 &   \mapsto {\rm ~NOT~prime} !
\\
\end{array}
$$

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection{As above but \texorpdfstring{$-1$ except $+1$: $p_1 \cdot p_2 \cdots p_n -1$}
                                        {-1 except +1: p\_1 ... p\_2 ... p\_n -1}}
$$
\begin{array}{lll}
2\cdot 3 -1     &   = 5 &   \mapsto {\rm ~prime} \\
2\cdot 3 \cdot  5  -1   &   = 29 &  \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 7  -1   &   = 11 \cdot 19 & \mapsto {\rm ~NOT~prime} ! \\
2\cdot 3 \cdots 11 -1   &   = 2,309 &    \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 13 -1   &   = 30,029 &    \mapsto {\rm ~prime} \\
2\cdot 3 \cdots 17 -1    &  = 61 \cdot 8,369 &   \mapsto {\rm ~NOT~prime!}
\end{array} 
$$

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection[Euclidean numbers \texorpdfstring{$e_n = e_0 \cdot e_1 \cdots e_{n-1} + 1$}
                                             {e\_n = e\_0 ... e\_1 e\_n-1 + 1}]
              {Euclidean numbers $e_n = e_0 \cdot e_1 \cdots e_{n-1} + 1$
	       with $ n \geq 1 $ and $ e_0 := 1 $}
%	       with $n$ greater than or equal to $1$ and $e_0 := 1$}
    \index{Euclidean number}
    $e_{n-1}$ is not the $(n-1)$th prime number, but the number 
    previously found here.
    Unfortunately this formula is not open but recursive.
    The sequence starts with

@@

% --------------------------------------------------------------------------
% \pagebreak %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vskip +10 pt
\subsection{\texorpdfstring{$f(n) = n^2 + n + 41$}{f(n) = n\^{}2 + n + 41}}
\label{L-Polynomfunktion01-41}
   This sequence starts off very {\em promisingly},  
   but is far from being a proof.

@@

% --------------------------------------------------------------------------
\pagebreak %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%   be_2005
% \vskip +10 pt
\subsection{\texorpdfstring{$f(n) = n^2 - 79 \cdot n + 1,601$}{f(n) = n\^{}2 - 79 ... n + 1,601}}
\label{L-Polynomfunktion02-1601}   \label{SrcPrimArith1}
   This function\footnote{%
   See \hyperlink{Primes_Appendix_A}{Appendix A of this chapter} for the source code
   to compute the table using Sage\index{Sage}.}  %Pari-GP.\index{Pari-GP}}
   delivers prime numbers for all values from $n=0$ to $n=79$.
Unfortunately $f(80) = 1,681 = 11 \cdot 151$ is not a prime number. To this
date, no function has been found that delivers more prime numbers in a row. On
the other hand, each prime occurs twice (first in the decreasing then in the
increasing sequence), which means that the algorithm delivers a total of 40
different prime values (these are the same ones as delivered by the function
in chapter~\ref{L-Polynomfunktion01-41})\footnote{%
   Another quadratic polynom, which delivers these primes, is: 
   $f(n) = n^2 - 9 \cdot n + 61$.\\
   Among the first 1000 sequence elements more than 50\% are prime
   (See \hyperlink{Primes_Appendix_A}{Appendix A of this chapter}).}.
$$
\begin{array}{|ll||ll|}
\hline
f(0)  = 1.601 & \mapsto {\rm ~prim}   &   f(26) =   223 & \mapsto {\rm ~prim} \\
f(1)  = 1.523 & \mapsto {\rm ~prim}   &   f(27) =   197 & \mapsto {\rm ~prim} \\
f(2)  = 1.447 & \mapsto {\rm ~prim}   &   f(28) =   173 & \mapsto {\rm ~prim} \\
f(3)  = 1.373 & \mapsto {\rm ~prim}   &   f(29) =   151 & \mapsto {\rm ~prim} \\
f(4)  = 1.301 & \mapsto {\rm ~prim}   &   f(30) =   131 & \mapsto {\rm ~prim} \\
f(5)  = 1.231 & \mapsto {\rm ~prim}   &   f(31) =   113 & \mapsto {\rm ~prim} \\
f(6)  = 1.163 & \mapsto {\rm ~prim}   &   f(32) =    97 & \mapsto {\rm ~prim} \\
f(7)  = 1.097 & \mapsto {\rm ~prim}   &   f(33) =    83 & \mapsto {\rm ~prim} \\
f(8)  = 1.033 & \mapsto {\rm ~prim}   &   f(34) =    71 & \mapsto {\rm ~prim} \\
f(9)  =   971 & \mapsto {\rm ~prim}   &   f(35) =    61 & \mapsto {\rm ~prim} \\
f(10) =   911 & \mapsto {\rm ~prim}   &   f(36) =    53 & \mapsto {\rm ~prim} \\
f(11) =   853 & \mapsto {\rm ~prim}   &   f(37) =    47 & \mapsto {\rm ~prim} \\
f(12) =   797 & \mapsto {\rm ~prim}   &   f(38) =    43 & \mapsto {\rm ~prim} \\
f(13) =   743 & \mapsto {\rm ~prim}   &   f(39) =    41 & \mapsto {\rm ~prim} \\
f(14) =   691 & \mapsto {\rm ~prim}   &   f(40) =    41 & \mapsto {\rm ~prim} \\              
f(15) =   641 & \mapsto {\rm ~prim}   &   f(41) =    43 & \mapsto {\rm ~prim} \\              
f(16) =   593 & \mapsto {\rm ~prim}   &   f(42) =    47 & \mapsto {\rm ~prim} \\              
f(17) =   547 & \mapsto {\rm ~prim}   &   f(43) =    53 & \mapsto {\rm ~prim} \\              
f(18) =   503 & \mapsto {\rm ~prim}   &   \cdots  &  \\                                       
f(19) =   461 & \mapsto {\rm ~prim}   &   f(77) = 1.447 & \mapsto {\rm ~prim} \\             
f(20) =   421 & \mapsto {\rm ~prim}   &   f(78) = 1.523 & \mapsto {\rm ~prim} \\             
f(21) =   383 & \mapsto {\rm ~prim}   &   f(79) = 1.601 & \mapsto {\rm ~prim} \\             
f(22) =   347 & \mapsto {\rm ~prim}   &   f(80) = 41 \cdot 41 & \mapsto {\rm ~NOT~prim!} \\
f(21) =   383 & \mapsto {\rm ~prim}   &   f(81) = 41 \cdot 43 & \mapsto {\rm ~NOT~prim!} \\ 
f(22) =   347 & \mapsto {\rm ~prim}   &   f(82) = 1.847 & \mapsto {\rm ~prim} \\             
f(23) =   313 & \mapsto {\rm ~prim}   &   f(83) = 1.933 & \mapsto {\rm ~prim} \\              
f(24) =   281 & \mapsto {\rm ~prim}   &   f(84) = 43 \cdot 47 &  \mapsto {\rm ~NOT~prim!} \\
f(25) =   251 & \mapsto {\rm ~prim}   &   & \\
\hline
\end{array} 
$$

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection[Polynomial functions 
    \texorpdfstring{$f(x) = a_n x^n + a_{n-1}x^{n-1} + \cdots + a_1 x^1 + a_0$}
                   {f(x) = a\_n x\^{}n + a\_n-1 x\^{}n-1 + ... + a\_1 x\^{}1 + a\_0}]
    {Polynomial functions\index{Polynomial} 
    $f(x) = a_n x^n + a_{n-1}x^{n-1} + \cdots + a_1 x^1 + a_0$ ~ 
    ($a_i$ in ${\mathbb Z}$, $n \geq 1$)}\index{Polynomial}

@@

     \begin{table}[ht]
     %\begin{center}
     \hspace*{1cm}\begin{tabular}{|r|r|r|r|l|}
     \hline
     Elements & First element & Distance & When & Discovered by \\ \hline
     22 &  11,410,337,850,553 &   4,609,098,694,200 & 1993 & Paul A. Pritchard,\\
        &                     &                     &      & Andrew Moran,\\
        &                     &                     &      & Anthony Thyssen\\
        &                     &                     &      &\\

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l||l|l||l|l|}\hline
\multicolumn{2}{|l||}{Ten-sized intervals} & \multicolumn{2}{l||}{Hundred-sized
intervals} & \multicolumn{2}{l|}{Thousand-sized intervals} \\ \hline
Interval  &     Number &    Interval  & Number &  Interval  & Number\\ \hline
\hline
1-10     &       4     &     1-100   &     25  &     1-1000     &    168 \\
11-20    &       4     &     101-200 &     21  &     1001-2000  &    135 \\
21-30    &       2     &     201-300 &     16  &     2001-3000  &    127  \\
31-40    &       2     &     301-400 &     16  &     3001-4000  &    120 \\
41-50    &       3     &     401-500 &     17  &     4001-5000  &    119 \\
51-60    &       2     &     501-600 &     14  &     5001-6000  &    114 \\
61-70    &       2     &     601-700 &     16  &     6001-7000  &    117 \\
71-80    &       3     &     701-800 &     14  &     7001-8000  &    107 \\
81-90    &       2     &     801-900 &     15  &     8001-9000  &    110 \\
91-100   &       1     &     901-1000 &     14 &      9001-10000 &    112 \\
\hline
\end{tabular}
\caption{How many primes exist within the first intervals of tens?}
\end{center}
\end{table}
\vskip +20 pt

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l|l|}\hline
Interval & Number & Average number per 1000 \\ \hline
1 - 10,000          &    1,229        &    122.900 \\
1 - 100,000         &    9,592        &    95.920 \\
1 - 1,000,000       &    78,498      &    78.498 \\
1 - 10,000,000       &   664,579     &    66.458 \\
1 - 100,000,000      &   5,761,455   &    57.615 \\
1 - 1,000,000,000    &   50,847,534  &    50.848 \\
1 - 10,000,000,000   &   455,052,512 &    45.505 \\ \hline
\end{tabular}
\caption{How many primes exist within the first intervals of dimensions?}
\end{center}
\end{table}
\vskip +6 pt

% --------------------------------------------------------------------------
\newpage
\hypertarget{ntePrimzahl}{}
\section{Appendix: Indexing prime numbers (\texorpdfstring{$n$}{n}-th prime number)} 
\label{s:ntePrimzahl}
\vskip +10 pt
